---
title: "Population Simulation with AlphaSimR: For *Streptanthus tortuosus* IPM Model"
author: "Rishav Ray"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This simulation models a plant population with multiple life history traits using AlphaSimR. The model includes size-dependent reproductive traits and exponential decay in QTL effect sizes to simulate realistic genetic architecture.

## Load Required Libraries

```{r libraries}
library(AlphaSimR)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(ggdoctheme)
library(dplyr)
library(patchwork)
library(gridExtra)
library(knitr)
library(rstatix)

# Set seed for reproducibility
set.seed(123)
```

# Simulation Parameters

```{r parameters}
# Population parameters
n_individuals = 1000
n_loci = 100
n_traits = 6

# Trait names
trait_names = c("Germination", "Initial_Size", "Establishment", 
 "Growth_Rate", "Flowering_Prob", "Fruit_Per_Plant")

trait_means = c(0.5, 2, 0.6, 1.2, 0.3, 2)  # Mean values for traits

# Heritabilities for each trait
h2 <- c(0.6, 0.5, 0.7, 0.4, 0.3, 0.4)
names(h2) <- trait_names

print("Simulation Parameters:")
print(paste("Number of individuals:", n_individuals))
print(paste("Number of loci per trait:", n_loci))
print(paste("Number of traits:", n_traits))
print("Heritabilities:")
print(h2)
```

# Genetic Architecture Setup

```{r genetic_architecture}
# Create founder genomes
founderPop = runMacs(nInd = 20,  # Small founder population
                      nChr = 10,   # 10 chromosomes
                      segSites = n_loci * n_traits / 10,  # Distribute loci across chromosomes
                      species = "GENERIC")

# Set up simulation parameters
SP = SimParam$new(founderPop)


# Add traits to simulation parameters

SP$addTraitA(nQtlPerChr = n_loci/10,  # Distribute across chromosomes
               mean = trait_means,
               var = c(1,1,1,1,1,1),
               gamma = TRUE,
               shape = 1,
               name = trait_names)

# Set heritabilities
SP$setVarE(h2 = h2)

print("Genetic architecture established with exponential decay in effect sizes")
```

# Generate F2 Population

```{r generate_f2}
# Create initial parents from founders
parents = newPop(founderPop, simParam = SP)

# Create F1 by random mating
F1 = randCross(parents, nCrosses = 500, simParam = SP)

# Create F2 population
F2 = randCross(F1, nCrosses = n_individuals, simParam = SP)

print(paste("Generated F2 population with", F2@nInd, "individuals"))
```

# Extract and Transform Trait Values

```{r extract_traits}

# Extract genetic values (breeding values) for all populations
gv_parents = parents@gv
gv_F1 = F1@gv
gv_F2 = F2@gv

# Extract phenotypes using @pheno slot
pheno_parents = parents@pheno
pheno_F1 = F1@pheno
pheno_F2 = F2@pheno
# Extract genetic values for all traits
#gv_matrix = F2@gv
#colnames(gv_matrix) = trait_names

# Set column names
colnames(gv_parents) = trait_names
colnames(gv_F1) = trait_names
colnames(gv_F2) = trait_names
colnames(pheno_parents) = trait_names
colnames(pheno_F1) = trait_names
colnames(pheno_F2) = trait_names

print("Genetic values and phenotypes extracted")
print("Dimensions check:")
print(paste("F2 genetic values:", dim(gv_F2)))
print(paste("F2 phenotypes:", dim(pheno_F2)))

# Convert to data frame
trait_data = as.data.frame(gv_F2)

# Transform traits to biologically meaningful scales
trait_data$Germination = plogis(trait_data$Germination * 0.5)  # Probability (0-1)
trait_data$Initial_Size = exp(trait_data$Initial_Size * 0.3 + 2)  # Size in mm
trait_data$Establishment = plogis(trait_data$Establishment * 0.4)  # Probability (0-1)
trait_data$Growth_Rate = exp(trait_data$Growth_Rate * 0.2 + 1)  # Growth multiplier
trait_data$Flowering_Prob_base = plogis(trait_data$Flowering_Prob * 0.3)  # Base probability
trait_data$Fruit_Per_Plant_base = exp(trait_data$Fruit_Per_Plant * 0.4 + 2)  # Base fruit number

# Add individual IDs
trait_data$Individual = 1:nrow(trait_data)

print("Trait transformation completed")
head(trait_data)
```

# Compare Genetic Values vs Phenotypes in F2

```{r compare_gv_pheno}
# Create comparison data frame for F2
comparison_data = data.frame(
  Individual = rep(1:nrow(gv_F2), n_traits),
  Trait = rep(trait_names, each = nrow(gv_F2)),
  Genetic_Value = as.vector(gv_F2),
  Phenotype = as.vector(pheno_F2)
)

# Calculate correlations between genetic values and phenotypes for each trait
correlations = sapply(1:n_traits, function(i) {
  cor(gv_F2[,i], pheno_F2[,i])
})
names(correlations) = trait_names

# Create summary table
correlation_summary = data.frame(
  Trait = trait_names,
  Heritability = h2,
  GV_Pheno_Correlation = correlations,
  Expected_Correlation = sqrt(h2)  # Theoretical expectation
)

kable(correlation_summary, digits = 3, 
      caption = "Genetic Value vs Phenotype Correlations")

print("Correlations between genetic values and phenotypes:")
print(round(correlations, 3))
```

# Visualization: Genetic Values vs Phenotypes

```{r gv_pheno_plots, fig.width=15, fig.height=10}
# Create scatter plots for each trait
plot_list = list()

for(i in 1:n_traits) {
  trait_data_all = data.frame(
    GV = gv_F2[,i],
    Phenotype = pheno_F2[,i]
  )
  
  p = ggplot(trait_data_all, aes(x = GV, y = Phenotype)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = paste(trait_names[i]),
         subtitle = paste("r =", round(correlations[i], 3), 
                         "| h² =", h2[i]),
         x = "Genetic Value (Breeding Value)",
         y = "Phenotype") +
    theme_minimal() +
    theme(plot.title = element_text(size = 12, hjust = 0.5))
  
  plot_list[[i]] = p
}

# Arrange plots
do.call(grid.arrange, c(plot_list, ncol = 3))
```

# Phenotype Distributions Across Generations

```{r phenotype_distributions}
# Prepare data for plotting distributions across generations

# Fix the trait assignment
dist_data <- data.frame(
  Generation = c(rep("Parents", nrow(pheno_parents) * n_traits),
                 rep("F1", nrow(pheno_F1) * n_traits),
                 rep("F2", nrow(pheno_F2) * n_traits)),
  Trait = c(rep(trait_names, nrow(pheno_parents)),
            rep(trait_names, nrow(pheno_F1)),
            rep(trait_names, nrow(pheno_F2))),
  Phenotype = c(as.vector(t(pheno_parents)),
                as.vector(t(pheno_F1)),
                as.vector(t(pheno_F2)))
)

# Set generation as factor with proper order
dist_data$Generation <- factor(dist_data$Generation, 
                              levels = c("Parents", "F1", "F2"))

print("Distribution data prepared")
print("Summary of phenotype distributions:")
summary_stats <- dist_data %>%
  group_by(Generation, Trait) %>%
  summarise(
    Mean = mean(Phenotype),
    SD = sd(Phenotype),
    Min = min(Phenotype),
    Max = max(Phenotype),
    .groups = "drop"
  )

kable(summary_stats, digits = 3, caption = "Phenotype Summary Statistics by Generation")
```


# Density Plots for Better Comparison

```{r density_plots, fig.width=15, fig.height=10}
# Create density plots for better comparison across generations
density_plots = list()

for(i in 1:n_traits) {
  trait_subset = dist_data[dist_data$Trait == trait_names[i], ]
  
  p = ggplot(trait_subset, aes(x = Phenotype, color = Generation, fill = Generation)) +
    geom_density(alpha = 0.3, size = 1) +
    scale_color_manual(values = c("Parents" = "red", "F1" = "green", "F2" = "blue")) +
    scale_fill_manual(values = c("Parents" = "red", "F1" = "green", "F2" = "blue")) +
    labs(title = paste("Phenotype Density:", trait_names[i]),
         x = "Phenotype Value",
         y = "Density") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  density_plots[[i]] = p
}

# Arrange density plots
do.call(grid.arrange, c(density_plots, ncol = 3))
```

# Variance Components Analysis

```{r variance_analysis}
# Calculate variance components for each trait in each generation
variance_analysis <- data.frame(
  Trait = rep(trait_names, 3),
  Generation = rep(c("Parents", "F1", "F2"), each = n_traits),
  Phenotypic_Variance = c(apply(pheno_parents, 2, var),
                         apply(pheno_F1, 2, var),
                         apply(pheno_F2, 2, var)),
  Genetic_Variance = c(apply(gv_parents, 2, var),
                      apply(gv_F1, 2, var),
                      apply(gv_F2, 2, var))
)

# Calculate environmental variance
variance_analysis$Environmental_Variance <- variance_analysis$Phenotypic_Variance - 
                                          variance_analysis$Genetic_Variance

# Calculate realized heritability
variance_analysis$Realized_Heritability <- variance_analysis$Genetic_Variance / 
                                          variance_analysis$Phenotypic_Variance

kable(variance_analysis, digits = 3, 
      caption = "Variance Components Analysis by Generation")

# Plot variance components
variance_long = variance_analysis %>%
  select(Trait, Generation, Genetic_Variance, Environmental_Variance) %>%
  pivot_longer(cols = c("Genetic_Variance", "Environmental_Variance"),
  names_to = "Variance_Component", 
       values_to = "Variance")

ggplot(variance_long, aes(x = Generation, y = Variance, fill = Variance_Component)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~Trait, scales = "free_y") +
  scale_fill_manual(values = c("Genetic_Variance" = "lightblue", 
                               "Environmental_Variance" = "lightcoral")) +
  labs(title = "Variance Components Across Generations",
       x = "Generation",
       y = "Variance",
       fill = "Variance Component") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Heritability Validation

```{r heritability_validation}
# Compare expected vs realized heritabilities
heritability_comparison <- data.frame(
  Trait = trait_names,
  Expected_h2 = h2,
  Realized_h2_Parents = variance_analysis$Realized_Heritability[1:n_traits],
  Realized_h2_F1 = variance_analysis$Realized_Heritability[(n_traits+1):(2*n_traits)],
  Realized_h2_F2 = variance_analysis$Realized_Heritability[(2*n_traits+1):(3*n_traits)]
)

kable(heritability_comparison, digits = 3, 
      caption = "Expected vs Realized Heritabilities")

# Plot heritability comparison
herit_long <- heritability_comparison |>
  pivot_longer(cols = -Trait, names_to = "Heritability_Type", values_to = "Heritability")

ggplot(herit_long, aes(x = Trait, y = Heritability, fill = Heritability_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Expected_h2" = "gold",
                               "Realized_h2_Parents" = "red",
                               "Realized_h2_F1" = "green", 
                               "Realized_h2_F2" = "blue")) +
  labs(title = "Expected vs Realized Heritabilities",
       x = "Trait",
       y = "Heritability",
       fill = "Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Size-Dependent Reproductive Traits

```{r size_dependent_traits}
# Calculate final size based on initial size and growth rate
trait_data$Final_Size = trait_data$Initial_Size * trait_data$Growth_Rate

# Size-dependent flowering probability
# Larger plants have higher flowering probability
size_effect_flowering = (trait_data$Final_Size - min(trait_data$Final_Size)) / 
                        (max(trait_data$Final_Size) - min(trait_data$Final_Size))

trait_data$Flowering_Prob = trait_data$Flowering_Prob_base * (0.5 + 0.5 * size_effect_flowering)
trait_data$Flowering_Prob = pmin(trait_data$Flowering_Prob, 0.95)  # Cap at 95%

# Size-dependent fruit production
# Only flowering plants produce fruit, and larger plants produce more
trait_data$Flowers = rbinom(nrow(trait_data), 1, trait_data$Flowering_Prob)

size_effect_fruit = (trait_data$Final_Size - min(trait_data$Final_Size)) / 
                    (max(trait_data$Final_Size) - min(trait_data$Final_Size))

trait_data$Fruit_Per_Plant = trait_data$Flowers * 
                             (trait_data$Fruit_Per_Plant_base * (0.3 + 0.7 * size_effect_fruit))

print("Size-dependent traits calculated")
print(paste("Proportion flowering:", round(mean(trait_data$Flowers), 3)))
print(paste("Mean fruits per flowering plant:", 
            round(mean(trait_data$Fruit_Per_Plant[trait_data$Flowers == 1]), 2)))
```

# Visualization

```{r visualizations, fig.width=12, fig.height=10}
# Create summary statistics
summary_stats = trait_data |>
get_summary_stats(type = "common")


# Print summary
kable(t(summary_stats), digits = 3, caption = "Trait Summary Statistics")

# Plot 1: Distribution of key traits
p1 = ggplot(trait_data, aes(x = Germination)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "skyblue") +
  labs(title = "Germination Rate Distribution", x = "Germination Probability", y = "Count") +
  theme_doc()

p2 = ggplot(trait_data, aes(x = Final_Size)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "lightgreen") +
  labs(title = "Final Size Distribution", x = "Final Size (mm)", y = "Count") +
  theme_doc()

p3 = ggplot(trait_data, aes(x = Flowering_Prob)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "orange") +
  labs(title = "Flowering Probability Distribution", x = "Flowering Probability", y = "Count") +
  theme_doc()

p4 = ggplot(trait_data, aes(x = Fruit_Per_Plant)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "pink") +
  labs(title = "Fruit Production Distribution", x = "Fruits per Plant", y = "Count") +
  theme_doc()

(p1 + p2)/( p3 + p4)
```

```{r size_relationships, fig.width=12, fig.height=8}
# Plot 2: Size-dependent relationships
p5 = ggplot(trait_data, aes(x = Final_Size, y = Flowering_Prob)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Size vs Flowering Probability", 
       x = "Final Size (mm)", y = "Flowering Probability") +
  theme_doc()

p6 = ggplot(trait_data[trait_data$Flowers == 1, ], 
             aes(x = Final_Size, y = Fruit_Per_Plant)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Size vs Fruit Production (Flowering Plants Only)", 
       x = "Final Size (mm)", y = "Fruits per Plant") +
  theme_doc()

p5 + p6
```

```{r trait_correlations, fig.width=10, fig.height=8}
# Plot 3: Trait correlations
cor_traits = trait_data |>
  select(Germination, Initial_Size, Establishment, Growth_Rate, 
         Final_Size, Flowering_Prob, Fruit_Per_Plant) |>
  cor()

# Create correlation heatmap
cor_df = as.data.frame(as.table(cor_traits))
names(cor_df) = c("Trait1", "Trait2", "Correlation")

ggplot(cor_df, aes(Trait1, Trait2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1)) +
  theme_doc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Trait Correlation Matrix")
```

# Fitness and Selection Analysis

```{r fitness_analysis}
# Calculate composite fitness measure
# Fitness = Germination × Establishment × (Fruit_Per_Plant + 1)
trait_data$Fitness = trait_data$Germination * 
                      trait_data$Establishment * 
                      (trait_data$Fruit_Per_Plant + 1)

# Standardize fitness
trait_data$Fitness_Std = scale(trait_data$Fitness)[,1]

# Selection analysis
selection_summary = trait_data |>
  summarise(
    Mean_Fitness = mean(Fitness),
    Var_Fitness = var(Fitness),
    CV_Fitness = sd(Fitness)/mean(Fitness),
    Prop_Reproductive = mean(Flowers),
    Mean_Reproductive_Success = mean(Fruit_Per_Plant[Flowers == 1])
  )

kable(selection_summary, digits = 3, caption = "Population Fitness Summary")

# Plot fitness distribution
ggplot(trait_data, aes(x = Fitness)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "purple") +
  geom_vline(xintercept = mean(trait_data$Fitness), color = "red", linetype = "dashed") +
  labs(title = "Fitness Distribution", 
       subtitle = paste("Mean fitness =", round(mean(trait_data$Fitness), 3)),
       x = "Composite Fitness", y = "Count") +
  theme_doc()
```

# Save Results

```{r save_results}
# Save the population data
#write.csv(trait_data, "data/F2_population_traits.csv", row.names = FALSE)

# Save genetic values matrix
#write.csv(gv_matrix, "data/F2_genetic_values.csv", row.names = FALSE)

print("Results saved to CSV files:")
print("- F2_population_traits.csv: All trait values and derived measures")
print("- F2_genetic_values.csv: Raw genetic values from AlphaSimR")
```

# Summary

In this simulation we successfully generated an F2 population of 1000 individuals with the following characteristics:

1. **Genetic Architecture**: Each trait controlled by 100 QTL with exponentially decaying effect sizes, creating realistic genetic architecture where few loci have large effects.

2. **Life History Traits**: 
   - Germination probability (0-1)
   - Initial and final size (cm)
   - Establishment probability (0-1)
   - Growth rate multiplier
   - Size-dependent flowering probability
   - Size-dependent fruit production

3. **Key assumptions**: 
   - Larger plants have higher flowering probability and fruit production
   - Only flowering plants produce fruit
   - Composite fitness incorporates multiple life stages
