---
title: "Population Simulation with AlphaSimR: For *Streptanthus tortuosus* IPM Model"
author: "Rishav Ray"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In this vignette, I would mostly be focusing on simulating few populations of *Strepthanthus tortuosus* and some phenotype values to feed them into the downstream IPM models.

## Load Required Libraries

```{r libraries}
library(AlphaSimR)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(ggdoctheme)
library(dplyr)
library(patchwork)
library(gridExtra)
library(knitr)
library(rstatix)
library(modelr)
library(broom)
library(rsample)
library(yardstick)

# Set seed for reproducibility
set.seed(123)
```

## Simulation Parameters

```{r parameters}
# Population parameters
n_individuals = 1000
n_loci = 100
n_traits = 6

# Trait names
trait_names = c("Germination", "Initial_Size", "Establishment", 
 "Growth_Rate", "Flowering_Prob", "Fruit_Per_Plant")

trait_means = c(0.5, 2, 0.6, 1.2, 0.3, 2)  # Mean values for traits

# Heritabilities for each trait
h2 = c(0.6, 0.5, 0.7, 0.4, 0.3, 0.4)
names(h2) = trait_names

print("Simulation Parameters:")
print(paste("Number of individuals:", n_individuals))
print(paste("Number of loci per trait:", n_loci))
print(paste("Number of traits:", n_traits))
print("Heritabilities:")
print(h2)
```

## Genetic Architecture Setup

```{r genetic_architecture}
# Create founder genomes
founderPop = runMacs(nInd = 50,  # Small founder population
                      nChr = 10,   # 10 chromosomes
                      segSites = n_loci * n_traits / 10,  # Distribute loci across chromosomes
                      species = "GENERIC")

# Set up simulation parameters
SP = SimParam$new(founderPop)


# Add traits to simulation parameters

SP$addTraitA(nQtlPerChr = n_loci/10,  # Distribute across chromosomes
               mean = trait_means,
               var = c(1,1,1,1,1,1),
               gamma = TRUE,
               shape = 1,
               name = trait_names)

# Set heritabilities
SP$setVarE(h2 = h2)

print("Genetic architecture established")

parents = newPop(founderPop, simParam = SP)
```

Now we extract and visualize the traits and genetic values

## Extract and Transform Trait Values

```{r extract_traits}

# Extract genetic values (breeding values) for all populations
gv_parents = parents@gv
meanG(parents)
meanP(parents)

# Extract phenotypes using @pheno slot
pheno_parents = parents@pheno


# Extract genetic values for all traits
plot_data = cbind(gv_parents, pheno_parents)
new_names = c("Germination_GV", "Initial_Size_GV", "Establishment_GV", 
            "Growth_Rate_GV", "Flowering_Prob_GV", "Fruit_Per_Plant_GV",
            "Germination_Pheno", "Initial_Size_Pheno", "Establishment_Pheno", 
            "Growth_Rate_Pheno", "Flowering_Prob_Pheno", "Fruit_Per_Plant_Pheno")
colnames(plot_data) = new_names

```


# Visualize Genetic Values and Phenotypes

```{r viz_traits}

ggplot(plot_data, aes(x = Germination_GV, y = Germination_Pheno)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Germination: Genetic Value vs Phenotype",
       x = "Genetic Value", y = "Phenotype") +
  theme_doc() +
  theme(plot.title = element_text(hjust = 0.5, size =14))
```

## Run a CV Analysis of prediction

```{r cv_analysis}
cv_results = tibble()
k = 5
repeats = 10

data = as.data.frame(plot_data) |>
mutate(Individual = row_number()) |>
dplyr::select(Individual,
    Genetic_Value = Germination_GV, 
             Phenotype = Germination_Pheno)


for (rep in 1:repeats) {
    cat(sprintf("Running CV repeat %d/%d\n", rep, repeats))
    
    # Create k-fold splits
    cv_folds = vfold_cv(data, v = k, strata = NULL)
    
    # Perform CV for this repeat
    fold_results = cv_folds |>
      mutate(
        # Fit model on training data
        model = map(splits, ~ lm(Phenotype ~ Genetic_Value, data = analysis(.x))),
        
        # Get model statistics on training data
        train_stats = map(model, glance),
        
        # Make predictions on test data
        predictions = map2(model, splits, ~ {
          test_data = assessment(.y)
          predictions = predict(.x, newdata = test_data)
          
          tibble(
            Individual = test_data$Individual,
            Actual = test_data$Phenotype,
            Predicted = predictions,
            Residual = Actual - Predicted
          )
        }),
        
        # Calculate fold-level metrics
        fold_metrics = map(predictions, ~ {
          tibble(
            RMSE = sqrt(mean(.x$Residual^2)),
            MAE = mean(abs(.x$Residual)),
            R_squared = cor(.x$Actual, .x$Predicted)^2,
            Correlation = cor(.x$Actual, .x$Predicted),
            Bias = mean(.x$Residual)
          )
        })
      )
    
    # Extract results for this repeat
    repeat_results = fold_results |>
      select(id, fold_metrics) |>
      unnest(fold_metrics) |>
      mutate(Repeat = rep)
    
    cv_results = bind_rows(cv_results, repeat_results)
}

cv_summary = cv_results |>
  group_by(Repeat) |>
  summarise(
    Mean_RMSE = mean(RMSE),
    Mean_MAE = mean(MAE),
    Mean_R_squared = mean(R_squared),
    Mean_Correlation = mean(Correlation),
    Mean_Bias = mean(Bias),
    .groups = "drop"
  )
```

# Display CV Summary
```{r display_cv_summary}
p1 = cv_summary |>
  ggplot(aes(x = factor(Repeat), y = Mean_R_squared)) +
  geom_boxplot(alpha = 0.7, fill = "lightblue") +
  geom_point() +
  geom_hline(yintercept = mean(cv_summary$Mean_R_squared), 
             color = "red", linetype = "dashed") +
  labs(title = "R² Across CV Repeats",
       x = "CV Repeat",
       y = "Mean R²") +
  theme_doc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.title = element_text(hjust = 0.5, size =14))

p2 = cv_summary |>
  ggplot(aes(x = factor(Repeat), y = Mean_RMSE)) +
  geom_boxplot(alpha = 0.7, fill = "lightcoral") +
  geom_point() +
  geom_hline(yintercept = mean(cv_summary$Mean_RMSE), 
             color = "red", linetype = "dashed") +
  labs(title = "RMSE Across CV Repeats",
       x = "CV Repeat", 
       y = "Mean RMSE") +
  theme_doc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.title = element_text(hjust = 0.5, size =14))

p3 = cv_summary |>
  ggplot(aes(x = factor(Repeat), y = Mean_Correlation)) +
  geom_boxplot(alpha = 0.7, fill = "lightgreen") +
  geom_point() +
  geom_hline(yintercept = mean(cv_summary$Mean_Correlation), 
             color = "red", linetype = "dashed") +
  labs(title = "Correlation Across CV Repeats",
       x = "CV Repeat",
       y = "Mean Correlation") +
  theme_doc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.title = element_text(hjust = 0.5, size =14))

p1
p2
p3
```

The correlation between the genetic values and phenotypes is quite high, indicating that the genetic architecture is well-defined and the traits are heritable. The CV analysis shows that the model performs consistently across different folds, with reasonable R² and RMSE values.

## How about we try and make a cross of two populations with different trait means?

```{r cross_populations}
trait_means_pop1 = c(0.5, 2, 0.6, 1.2, 0.3, 2)  # Mean values for traits
trait_means_pop2 = c(0.9, 1.2, 1, 1.8, 0.7, 1.3)  # Mean values for traits

# Heritabilities for each trait
h2 = c(0.6, 0.5, 0.7, 0.4, 0.3, 0.4)
names(h2) = trait_names

founderPop = runMacs(nInd = 50,  # Small founder population
nChr = 10,   # 10 chromosomes
segSites = n_loci * n_traits / 10,  # Distribute loci across chromosomes
species = "GENERIC")

# Set up simulation parameters
SP1 = SimParam$new(founderPop)
SP2 = SimParam$new(founderPop)


# Add traits to simulation parameters

SP1$addTraitA(nQtlPerChr = n_loci/10,  # Distribute across chromosomes
               mean = trait_means_pop1,
               var = c(1,1,1,1,1,1),
               gamma = TRUE,
               shape = 1,
               name = trait_names)

SP2$addTraitA(nQtlPerChr = n_loci/10,  # Distribute across chromosomes
               mean = trait_means_pop2,
               var = c(1,1,1,1,1,1),
               gamma = TRUE,
               shape = 1,
               name = trait_names)

# Set heritabilities
SP1$setVarE(h2 = h2)
SP2$setVarE(h2 = h2)

print("Genetic architecture established")

pop1 = newPop(founderPop, simParam = SP1)
pop2 = newPop(founderPop, simParam = SP2)

mega_pop = newMultiPop(pop1, pop2)


```

```{r cross_pops}

# Now we can cross the two pops
randCross = randCross2(pop1, pop2, nCrosses = 1000)

# Extract genetic values and phenotypes for the crossed population
gv_crossed = randCross@gv
pheno_crossed = randCross@pheno
# Combine genetic values and phenotypes into a single data frame
crossed_data = cbind(gv_crossed, pheno_crossed)
new_names_crossed = c("Germination_GV", "Initial_Size_GV",
                        "Establishment_GV", "Growth_Rate_GV",
                        "Flowering_Prob_GV", "Fruit_Per_Plant_GV",
                        "Germination_Pheno", "Initial_Size_Pheno",
                        "Establishment_Pheno", "Growth_Rate_Pheno",
                        "Flowering_Prob_Pheno", "Fruit_Per_Plant_Pheno")
colnames(crossed_data) = new_names_crossed
# Visualize the genetic values and phenotypes for the crossed population
ggplot(crossed_data, aes(x = Germination_GV, y = Germination_Pheno)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Germination: Genetic Value vs Phenotype (Crossed Population)",
       x = "Genetic Value", y = "Phenotype") +
  theme_doc() +
  theme(plot.title = element_text(hjust = 0.5, size =14))

# Visualize the distribution of genetic values and phenotypes for the crossed population
ggplot(crossed_data, aes(x = Germination_Pheno)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Germination Genetic Values (Crossed Population)",
       x = "Genetic Value", y = "Count") +
  theme_doc() +
  theme(plot.title = element_text(hjust = 0.5, size = 14))

meanP(randCross)
meanG(randCross)
```