---
title: "IPM genetic simulation"
author: Rishav Ray
output: 
 - html_document
 - md_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

This workbook is mainly for simulating the genetic data for the IPM project. The simulation is done using the AlphaSimR package.

```{r, message = FALSE, warning = FALSE}
library(AlphaSimR)
library(tidyverse)
library(ggthemes)
library(ggdoctheme)
```

# Create founder population

```{r}
founderPop = runMacs(
  nInd = 100,
  nChr = 10,
  segSites = 1000,
  species = "GENERIC"
)
```


# Define a simple genetic architecture for a trait influenced by environment

```{r}
SP = SimParam$new(founderPop)
SP$addTraitAG(nQtlPerChr = 20, mean = 5, var = 2, varGxE =0.2)
VarE = 0.4
SP$setVarE(h2 = 0.5) # Heritability of 0.5
```


# Create initial population
```{r}
pop = newPop(founderPop, simParam = SP)
```


# Add environmental effect - simulate climate 
Currently not running this, but we can use this function later to customize environmental interaction
```{r}
environmentalEffect = function(pop, climateVar) {
  n = pop@nInd
  # Climate effect proportional to individual's genetic value for adaptation
  effect = pop@gv * climateVar
  # Add environmental noise
  effect = effect + rnorm(n, 0, 0.5)
  return(effect)
}
```

# Simulate multiple generations with changing environment
We would be using the setPheno function to set the phenotypic values with the residual environmental variance.
```{r}
results = list()
climateTrend = seq(1, 1.5, length.out = 10) # Warming climate scenario

for (gen in 1:10) {
  # Current climate variable
  currentClimate = climateTrend[gen]
  
  # Calculate fitness based on genetic value and environment
  pop = setPheno(pop, varE = VarE*currentClimate, reps = 1, simParam = SP)
  fitness = pop@pheno[,1]
  
  # Selection - individuals with higher fitness have higher probability of reproducing
  selectedIndices = sample(1:pop@nInd, 50, prob = fitness/sum(fitness))
  selectedPop = pop[selectedIndices]
  
  # Random mating
  pop = randCross(selectedPop, nCrosses = 100, nProgeny = 1, simParam = SP)
  
  # Store results
  results[[gen]] = data.frame(
    generation = gen,
    meanGV = mean(pop@gv),
    varGV = var(pop@gv[,1]),
    climate = currentClimate,
    fitness = mean(fitness)
  )
}
```

# Visualize results

```{r}
# Convert results to data frame
resultsDF = do.call(rbind, results)

ggplot(resultsDF, aes(x = generation)) +
  geom_line(aes(y = meanGV, color = "Mean Genetic Value")) +
  geom_line(aes(y = climate, color = "Climate")) +
  geom_line(aes(y = varGV, colour = "Trait Variance"))+
  geom_line(aes(y =fitness, colour = "Fitness"))+
  labs(title = "Adaptation to Changing Climate",
       x = "Generation",
       y = "Value") +
  #scale_color_manual(values = c("Mean Genetic Value" = "blue", "Climate" = "red")) +
  theme_doc()
```

