---
title: "IPM WL2 garden with climate"
author: Rishav Ray
output: 
 - md_document
 - html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# Data load and wrangling the temperature and moisture data to the main data

First load some libraries
```{r, message=FALSE}
library(tidyverse)
library(ggthemes)
library(ggdoctheme)
library(lme4)
library(lmerTest)
library(ggpubr)
library(ggrepel)
library(patchwork)
library(lubridate)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
select = dplyr::select
```

Now load the data

```{r}
data = read_tsv("data/survival_data.tsv")
temp_data = read_csv("data/WL2_2022_2023_iButton_Data_Corrected.csv") |>
mutate(datetime_obj = mdy_hm(Date_Time)) |>
mutate(date = as_date(datetime_obj), time = format(datetime_obj, "%H:%M")) |>
filter(date > as.Date("2023-07-06")) |>
select(Bed, date, time, SoilTemp,)

moisture_data = read_csv("data/WL2_2023_Bed_C_Soil_Moisture_Corrected.csv") |>
mutate(datetime_obj = mdy_hm(Date_Time)) |>
mutate(date = as_date(datetime_obj), time = format(datetime_obj, "%H:%M")) |>
filter(date > as.Date("2023-07-06")) |>
select(date, time, Port_1:Port_5)

```

Take the mean of the 5 ports from the moisture sensor data and summarize to daily with max and min

```{r}
mean_moisture = moisture_data |>
rowwise() |>
mutate(moisture_avg = mean(Port_1:Port_5, na.rm = TRUE)) |>
select(date, time, moisture_avg)


daily_moisture = mean_moisture |>
ungroup() |>
group_by(date) |>
summarize(moisture_max = max(moisture_avg, na.rm = TRUE), moisture_min = min(moisture_avg, na.rm = TRUE), 
mean_moisture = mean(moisture_avg, na.rm = TRUE), moisture_cv = sd(moisture_avg, na.rm = TRUE)/mean(moisture_avg, na.rm = TRUE)*100)
```

Make some plots to see the CV over time

```{r}
daily_moisture |>
ggplot(aes(x = date, y = moisture_cv, group = 1)) +
geom_line() +
geom_point() +
theme_doc()+
labs(x = "Date", y = "CV of moisture (%)")
```

Perform the same with soil temperature data

```{r}
daily_temp = temp_data |>
group_by(Bed, date) |>
summarize(temp_max = max(SoilTemp, na.rm = TRUE), temp_min = min(SoilTemp, na.rm = TRUE),
mean_temp = mean(SoilTemp, na.rm = TRUE), temp_cv = sd(SoilTemp, na.rm = TRUE)/mean(SoilTemp, na.rm = TRUE)*100)
```

Visualize the CV by bed

```{r}
daily_temp |>
ggplot(aes(x = date, y = temp_cv, group = Bed, color = Bed)) +
geom_line() +
geom_point() +
theme_doc()+
labs(x = "Date", y = "CV of temperature (%)")
```

Join the temperature and moisture data together to make things easier. Also, I am going to take the mean of all the beds

```{r}
daily_temp_moisture = left_join(daily_temp |> 
ungroup() |> 
group_by(date) |> 
summarize(mean_temp = mean(mean_temp, na.rm = TRUE),
mean_cv = mean(temp_cv, na.rm = TRUE),
temp_max = max(temp_max, na.rm = TRUE), temp_min = min(temp_min)),
daily_moisture, by = "date")
```

Now let's join the two data, by aggregating the daily data to weekly by anchoring it each day
Already ran, so just load it.
```
size_data_with_intervals = data |>
    arrange(Genotype, survey_date) |>
    group_by(Genotype) |>
    mutate(
      prev_survey_date = lag(survey_date),
      # If it's the first measurement for a Genotype, use 7 days as default lookback
      lookback_start = if_else(is.na(prev_survey_date), 
                               survey_date - days(7), 
                               prev_survey_date + days(1)),
      lookback_end = survey_date
    ) |>
    ungroup()
  
# Join the data with intervals

combined_data = size_data_with_intervals |>
    rowwise() |>
    mutate(
      # Filter environmental data for the specific interval
      matching_env_data = list(
        daily_temp_moisture |>
          filter(date >= lookback_start, date <= lookback_end)
      ),
      # Count how many env observations we have for this period
      env_obs_count = nrow(matching_env_data),
      # Calculate aggregates
      period_temp_mean = mean(matching_env_data$mean_temp, na.rm = TRUE),
      period_temp_cv = mean(matching_env_data$mean_cv, na.rm = TRUE),
      period_temp_max = max(matching_env_data$temp_max, na.rm = TRUE),
      period_temp_min = min(matching_env_data$temp_min, na.rm = TRUE),
      period_moisture_max = max(matching_env_data$moisture_max, na.rm = TRUE),
      period_moisture_min = min(matching_env_data$moisture_min, na.rm = TRUE),
      period_moisture_mean = mean(matching_env_data$mean_moisture, na.rm = TRUE),
      period_moisture_cv = mean(matching_env_data$moisture_cv, na.rm = TRUE)
    ) |>
    # Clean up intermediate columns
    select(-matching_env_data) |>
    ungroup()

```

```{r}
#write_tsv(combined_data, "data/combined_data.tsv")
combined_data = read_tsv("data/combined_data.tsv")
```

Now visualize the mean temperature vs size by population

```{r}
ggplot(combined_data, aes(x = period_temp_mean, y = size)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ population) +
  theme_doc() +
  labs(title = "Effect of Mean Temperature on Plant Size",
       x = "Mean Temperature (°C)", y = "Plant Size")
```

Heatmap of growth rate vs environmental variables

```{r}
# First calculate growth rate
grate = combined_data |>
  group_by(Genotype) |>
  mutate(growth_rate = (sizeNext - size)/7) |>
  ungroup()
  
# Then plot heatmap
ggplot(grate, aes(x = period_temp_mean, y = period_moisture_max, color = growth_rate)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  theme_doc()+
  labs(title = "Growth Rate by Temperature and Moisture",
       x = "Mean Temperature (°C)", y = "Maximum Moisture")
```

```{r}
# Create long format for environmental variables
env_long = combined_data |>
  pivot_longer(
    cols = c(period_temp_mean, period_temp_cv, period_moisture_mean, period_moisture_cv),
    names_to = "env_variable",
    values_to = "env_value"
  )
  
ggplot(env_long, aes(x = env_value, y = sizeNext)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  facet_wrap(~ env_variable, scales = "free_x") +
  theme_doc()+
  labs(title = "Size Response to Different Environmental Variables")
```

```{r}
ggplot(combined_data, aes(x = size, y = sizeNext, color = period_temp_mean)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_viridis_c(option = "plasma") +
  theme_doc()+
  labs(title = "Growth Transitions Colored by Temperature",
       x = "Current Size", y = "Next Size", colour = "Mean Temperature (°C)")
```

```{r}
ggplot(combined_data, aes(x = size, y = sizeNext, color = period_moisture_mean)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_viridis_c(option = "plasma") +
  theme_doc()+
  labs(title = "Growth Transitions Colored by Moisture %",
       x = "Current Size", y = "Next Size", colour = "Mean Moisture (%)")
```

library(lme4)

```{r}
# Fit a simple mixed model
model = lmer(sizeNext ~ period_temp_cv + period_moisture_cv + (1|population), data = combined_data)
summary(model)
# Extract and plot predictions
plotdata = combined_data |>
select(sizeNext, period_temp_mean, period_moisture_max, population) |>
na.omit() |>
mutate(predicted = predict(model))

ggplot(plotdata, aes(x = predicted, y = sizeNext, color = population)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  theme_doc()+
  labs(title = "Mixed Effects Model Predictions vs Observed Size",
       x = "Predicted Size", y = "Observed Size")
```

```{r}
# Check the survival model
model = glmer(survival ~ size + period_temp_cv + period_moisture_mean + (size | population), family = binomial(), data = combined_data)
summary(model)
# Extract and plot predictions
plotdata = combined_data |>
select(survival, size, period_temp_cv, period_moisture_mean, population) |>
na.omit() |>
mutate(pred_survival = plogis(predict(model)))

plotdata |>
ggplot(aes(x = factor(survival), y = pred_survival)) +
geom_boxplot(outlier.shape = NA)+
geom_jitter(width = 0.1, height = 0.1, alpha = 0.5) +
labs(x = "Observed Survival", y = "Predicted Survival",
     title = "Observed vs Predicted Survival") +
theme_doc()
```

# Let's make an environment dependent kernel

```{r}
fit_hierarchical_growth = function(data) {
  # Fit hierarchical growth model using lme4
  growth_model = lmer(sizeNext ~ size + poly(size,2) + period_temp_cv + period_moisture_cv + (size | population) + (1|survey_date),
    data = data
)
  
  return(growth_model)
}

# Function to fit hierarchical survival model
fit_hierarchical_survival = function(data) {
  # Fit hierarchical survival model
  survival_model = glmer(
    survival ~ size + poly(size,3) + period_temp_cv + period_moisture_mean + (size | population),
    family = binomial(),
    data = data
  )
  
  return(survival_model)
}
```

Check model with and without climate data
```{r}
growth_model1 = lmer(sizeNext ~ size + period_temp_cv + period_moisture_cv + (size | population) + (1|survey_date),
    data = combined_data)
growth_model2 = lmer(sizeNext ~ size + period_moisture_cv + (size | population) + (1|survey_date),
    data = combined_data)
summary(growth_model1)
summary(growth_model2)
anova(growth_model1, growth_model2)

```
# Make kernel
```
# Define the integration range and mesh points
min_size <- min(result$size, na.rm = TRUE)
max_size <- max(result$size, na.rm = TRUE)
n_mesh <- 100
mesh_points <- seq(min_size, max_size, length.out = n_mesh)

# Environment-dependent kernel function
make_kernel <- function(env_temp, env_moisture) {
  
  # Get fitted vital rate models
  growth <- growth_model(mesh_points, env_temp, env_moisture)
  survival <- survival_model(mesh_points, env_temp, env_moisture)
  
  # Initialize kernel matrix
  K <- matrix(0, n_mesh, n_mesh)
  
  # Build the kernel
  for (i in 1:n_mesh) {
    for (j in 1:n_mesh) {
      z <- mesh_points[j]  # Current size
      zp <- mesh_points[i] # Next size
      
      # Growth kernel component
      mean_size_next <- growth$mean_function(z, env_temp, env_moisture)
      variance <- growth$variance_function(z)
      g_z_zp <- dnorm(zp, mean = mean_size_next, sd = sqrt(variance))
      
      # Survival component
      s_z <- survival$function(z, env_temp, env_moisture)
      
      # Combine into the kernel
      K[i, j] <- s_z * g_z_zp
    }
  }
  
  # Normalize the kernel to ensure proper integration
  h <- (max_size - min_size) / (n_mesh - 1)  # Step size
  K <- K * h
  
  return(K)
}

```